A second alternative to the pairwise testing approach

### 2022-08-29 text/22/analysis-of-means.Rmd
Good first draft of analysis-of-means.Rmd

### Comparison to an overall mean

Suppose you are in a setting like the one described above, where there is an expectation of similarity from one r goal is to establish whether any groups differ from the overall mean, Then you would write your hypothesis as

```{r, fig.width=5, fig.height=0.33, echo=FALSE}
h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu ~ " for all i=1,...,k")
par(mar=rep(0.1, 4))
plot(0:1, 0:1, axes=FALSE, type="n")
text(0.1, 0.5, h0, adj=0)
```

This approach is not that common, but you can find it in most statistical software programs under the name "Analysis of Means." It fits in well with a setting where you hope that all the groups produce consistent results. If any do not, then you want to identify the group or groups that deviate from the norm and study them further.

@@

@@

The raw data was not available, but you can calculate the analysis of means limits using the group means and standard deviations.

```{r}
mn <- c(
  24.87,
  24.54,
  23.61,
  18.32,
  22.72,
  22.81)

sd <- c(
  0.56,
  1.14,
  1.33,
  2.45,
  0.19,
  1.18)
mse <- mean(sd^2)
mse
ybarbar <- mean(mn)
t <- 5
N <- 20
h <- 2.88
hi <- ybarbar+h*sqrt((t-1)*mse/N)
lo <- ybarbar-h*sqrt((t-1)*mse/N)
lo
hi
```

https://www.frontiersin.org/articles/10.3389/fchem.2022.894547/full

>>>>

### The comparison to the overall mean hypothesis

Suppose you are in a setting like the one described above, where there is an expectation of similarity between each group mean and the overall mean. Then you would write your hypothesis as 
```{r, fig.width=5, fig.height=0.33, echo=FALSE}
h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu ~ " for all i=1,...,k")
par(mar=rep(0.1, 4))
plot(0:1, 0:1, axes=FALSE, type="n")
text(0.1, 0.5, h0, adj=0)
```

This approach is not that common, but you can find it in most statistical software programs under the name "Analysis of Means." It fits in well with a setting where you hope that all the groups produce consistent results. If any do not, then you want to identify the group or groups that deviate from the norm and study factors that may account for the deviation.

The analysis of means approach that compares each group mean to the overall mean is easy to implement and it lends itself to a simple graphical display.

You don't need to compute the traditional F statistic for Analysis of Variance first, because the Analysis of Means approach controls the overall Type I error rate. This protects you from the accusation of p-hacking, even if the number of groups is very large.

@@

![](http://www.pmean.com/new-images/22/analysis-of-means-00.png)

(Image taken from Wikipedia)

@@

https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0157498

Here are the summary statistics.

```{r, echo=FALSE}
mn <- c(
  21.83,
  20.40,
  21.62,
  21.06,
  20.57)

sd <- c(
  1.17,
  1.09,
  1.32,
  1.11,
  0.99)

mse <- mean(sd^2)
ybarbar <- mean(mn)
t <- 5
N <- 20
h <- 2.88
hi <- ybarbar+h*sqrt((t-1)*mse/N)
lo <- ybarbar-h*sqrt((t-1)*mse/N)
```

![](http://www.pmean.com/new-images/22/analysis-of-means-01.png)

The raw data was not available, but you can calculate the analysis of means limits using the group means and standard deviations. You can plot the individual means versus the limits provided by analysis of means calculations.

![](http://www.pmean.com/new-images/22/analysis-of-means-02.png)

Notice that all five means lie inside the limits. None of the five strains shows a  statisticially significant difference from the overall mean.

### Caveats

The formulas get a bit more complex and the graph gets a bit messier if the sample size differs from one group to another.

More importantly, the only comparisons that you can make with the analysis of means approach is a comparison to the overall mean. If two groups are both statistically significantly higher than the overall mean, you cannnot make a comparison between those two groups without losing control over the overall Type I error rate.

The Analysis of Means, just like the Dunnett's test described in an earlier blog post, provides a simple approach to testing a different type of hypothesis. Because it reduces the number of comparisons from all possible pairwise differences to a comparison of each group to the overall mean, you gain some precision and can summarize your results in a simple easily understood graph.
